\section{The \re{187}-\os{187} chronometer}
In this work, \textit{cosmochronology} refers to dating astrophysical events based, more specifically the age of nucleosynthesis.
A \textit{chronometer} is a pair of radioactive nuclei, with an appropriate halflife which allows us to do exactly that.
As mentioned in section \comment{section nuclear physics with betadecay \ref{sec:betadecay}?}, \re{187} \betadecay{s} to \os{187} with a halflife of 41.6 Gyr (\mycite{snelling15}).
This halflife is appropriate for dating Galactic ages, which occur on the scale of universal ages (The age of the universe, from the Planck collaboration is measured to 13.8 Gyrs \mycite{planck15}).
Because of this, \mycite{clayton64} suggest the \re{187}-\os{187} nuclei as Galactic chronometer, also called a cosmic clock.

\comment{insert tikz chart of nuclides here \\}
From section \comment{nuclear-physics r-process and s-process} it is clear that synthesis of heavy nuclei can be divided into two processes, rapid and slow neutron capture (ignoring proton capture).
Figure \comment{tikz chart of nuclides} show the s-process path and the r-process path in the region around the \re{187}-\os{187}-pair.

\os{187} and \os{187} can only be reached by the s-process, because the r-process material will \betadecay to \re{187} and \w{186} where the halflife is comparable to the age of the universe. Simontaneously, neglecting the \w{184} branching points, the s-process cannot reach \re{187} (\mycite{clayton64}).

This means that \re{187} is synthesized through the r-process and \os{187} is synthesized through the s-process from \os{186} and cosmoradiogenic \betadecay from \re{187}.
By assuming a rate of events producing r-process elements, starting from the age of nucleosynthesis, the amount of cosmoradiogenic \os{187} compared to \re{187} in the interstellar medium Galaxy is calculatable.
This ratio is assumed to be representative of the interstellar medium near the \sos during it's formation.
The same ratio can be found at the formation of the \sos from meteorites today. These meteorites condensed from the dust and gas in the early epoch of the \sos and have remained pure since then until they impacted the earth.

In his model for cosmochronology of Re-Os, \mycite{clayton64} assumes a exponential decay of the rate of r-process events $A(t) = e^{-\Lambda t}$, where an infinite decay constant ($\Lambda\rightarrow\infty$) leads to a ``sudden synthesis'' model and a zero decay constant ($\Lambda\rightarrow 0$) leads to a ``uniform synthesis'' model.
A detailed calculation following these models are presented in appendx \ref{sec:calc-cosmo-chronology}.


\iffalse

\comment{include chart of nuclide section}
%% \begin{figure}
%%   \input{../tikzfigures/nuclides187_chart}
%% \end{figure}


From the chart on can see the usual path for slow neutron capture along the valley of stability.
This is the main contribution to \os{187}. In a standard s-process analysis, \w{185} and \re{186} are unstable, and will decay before they can capture neutrons. The s-process path will never synthesize \re{187}. However if the neutron capture rate is comparable to the \betadecay rates of those nuclides a branching point can occur. A branching point is a point where the synthesizing path of the s-process split due to competing nuclear reactions. In this case a significant fraction of the s-process path can go through \re{186} to \re{187} or through \w{185} to \w{186} (which is stable) and onwards to \re{187}. Apart from these effects \re{187} is shielded from s-process contribution.

The rapid neutron capture maintains very high neutron numbers until the neutron source ``shuts off'', at that point the isotopes \betadecay to the valley of stability. Given the long \halflife of \re{187} it can be considered stable, so the \os{187} isotope is shielded from r-process contribution because almost all \betadecay on the 187-isobar will stop at \re{187}.

Since \re{187} is radioactive with a halflife of \comment{add here} some \re{187} in the interstellar or stellar medium will decay to \os{187}. This amount is called the cosmoradiogenic \os{187}. The fraction between cosmoradiogenic \os{187} and current \re{187} is given by the exponential decay-function (assuming ofcourse that the nuclear decay rate is constant for all time, including stellar environments), meaning the time of nucleosynthesis can be calculated from the observed fraction of daughter-parent nuclei.

Clayton attempts to calculate the fraction of cosmoradiogenic osmium and the age of nucleosynthesis from these principles\mycite{clayton64}.
A brief summary follows:

The abundance of \os{186} is due to s-process only, and the abundance of \os{187} is due to s-process (from \os{186}) and cosmo radiogenic enrichment from \re{187} beta-decay.
The s-process contribution from \os{186} is given by the \os{186} abundance and the ratio between the cross-section of the isotopes. It is shown from nebular Samarium that the two s-only isotopes have nearly identical cross-section times abundance. Extrapolating this to other s-process isotopes, the result is:
(denoting abundance of isotope by their chemical name instead of N for simplicity)
\begin{equation}
  \bar{\sigma}_{\os{186}}{\os{186}} = \bar{\sigma}_{\os{187}}{\os{186}_S}
  \quad \rightarrow \quad
  {\os{186}_S} = \frac{\bar{\sigma}_{\os{186}}}{\bar{\sigma}_{\os{187}}}{\os{186}}
\end{equation}
$\bar{\sigma}$ are the neutron-capture cross-sections averaged over the appropriate thermal velocity distributions.
The cosmoradiogenic component of \os{187} is therefor the remaining part.
\begin{equation}
  \os{187}_c = \os{187} - \os{187}_s = \os{187} - \frac{\bar{\sigma}_{\os{186}}}{\bar{\sigma}_{\os{187}}}{\os{186}}
\end{equation}
Rewriting equation to relative units.
\begin{equation}
  \frac{\os{187}_c}{\re{187}} = \frac{\os{187} - \frac{\bar{\sigma}_{\os{186}}}{\bar{\sigma}_{\os{187}}}{\os{186}}}{\re{187}} =
  \frac{
    \left(\frac{\os{187}}{Os}\right) -
    \frac{\bar{\sigma}_{\os{186}}}{\bar{\sigma}_{\os{187}}}
    \left(\frac{\os{186}}{Os}\right)
  }{
    \left(\frac{\re{187}}{Re}\right)
  }
  \left(\frac{Os}{Re}\right)
\end{equation}
Origin of time, t=0, is set to \sos formation. Assuming that r-process events are supernovae and they began occuring at time T before \sos formation(t=0), and the frequency of events decreases exponentially as $f_0e^{\Lambda t}$, where $f_0$ is the initial supernovae freqency.
According to this model, and Clayton, the amount of cosmoradiogenic \os{187} is given by
\begin{equation}
  \frac{\os{187}_c}{\re{187}} = \left[
    \frac{\Lambda-\lambda}{\Lambda}
    e^{\lambda T}
    \frac{1-e^{-\Lambda T}}{1-e^{-(\Lambda-\lambda) T}}
    \right] - 1
\end{equation}
And the two special cases. \\
Sudden synthesis ($\Lambda\rightarrow\infty$)
\begin{equation}
  \frac{\os{187}_c}{\re{187}} = e^{\lambda t} - 1
\end{equation}
Uniform synthesis ($\Lambda\rightarrow 0$)
\begin{equation}
  \frac{\os{187}_c}{\re{187}} = \frac{\lambda T}{1-e^{-\lambda t}} - 1
\end{equation}

\comment{end of direct summary}

In short, by modelling r-process nucleosynthesis and s-process nucleosynthesis, the fraction of \re{187}-\os{187} in the \sos can predict the time of nucleosynthesis from non-cosmological methods.
\comment{\\ What do I mean about the age of nucleosynthesis?}
\fi
